High-resolution ALMA observations of V4046\,Sgr: a circumbinary disc with a thin ring

### 2021-05-27 Draft_V4046Sgr_mnras.tex
Updates from Overleaf

% Abstract of the paper
\begin{abstract}
    The nearby V4046\,Sgr spectroscopic binary hosts a gas-rich disc known for its wide cavity and dusty ring. We present high resolution ($\sim$20--50\,mas) ALMA observations of the 1.3\,mm continuum of V4046\,Sgr which, combined with SPHERE--IRDIS polarized images and well-sampled spectral energy distribution (SED), allow us to propose a physical model using radiative transfer (RT) predictions. The new ALMA data reveal a impressively thin ring at a radius of 13.15$\pm$0.42\,au (Ring13), with a marginally resolved radial width of 2.46$\pm$0.20\,au. Ring13 is surrounded by a $\sim$10\,au-wide gap, and it is flanked by a mm-bright outer ring (Ring24) with a sharp inner edge at 24\,au. Between 25 and $\sim$35\,au the brightness of Ring24 is relatively flat and then breaks into a steeper tail that reaches out to $\sim$60\,au. The observation also detects a central emission that we interpret as a tight ring made of dust grains with a lower size limit of 0.5\,mm. In order to reproduce the SED, the model that we propose requires an inner ring at $\sim$5\,au (Ring5) composed mainly of small dust grains, hiding under the IRDIS coronagraph, and surrounding an inner circumbinary disc. The surprisingly thin Ring13 is nonetheless roughly 10 times wider than its expected vertical extent. The strong near-far disc asymmetry at 1.65\,$\micron$ points at a very forward-scattering phase function, and requires grain radii of no less than 0.4\,$\micron$. The previously reported scattered-light shadow of the secondary star is also reproduced by our physical model.
\end{abstract}

@@

Recent observations of young circumstellar discs have transformed the current knowledge on planet formation. The focus of resolved imaging with the Atacama Large Millimeter/submillimeter Array (ALMA) or with the current generation of high-contrast cameras has mainly been towards the brighter sources \citep[e.g.][]{2020ARA&A..58..483A}. It is to address this bias that the programme called ``Discs Around T Tauri Stars with SPHERE'' (DARTTS-S) collected differential polarization imaging (DPI) data secured with the Spectro-Polarimeter High-contrast Exoplanet REsearch \citep[SPHERE][]{2019A&A...631A.155B} in a total of 29 solar-type stars \citep[][]{Avenhaus_2018,Garufi2020}. The sample is not biased towards exceptionally bright and large discs. The DARTTS observations revealed diverse structures and morphologies in the scattering surface of these discs. This article on V4046 Sagittarii (Sgr) is the first installment of a companion programme, the DARTTS survey with ALMA (DARTTS-A), which will present millimetre observations of nine protoplanetary discs previously imaged in polarized scattered light in DARTTS-S.

Disc structures can be explored by means of comparing millimetre continuum and scattered-light images. These observations allow us to see the distribution of different populations of dust particles: while ALMA traces the millimetre-sized grains settled in the mid-plane of the disc, scattered-light imaging accounts for photons reflected off micron-sized dust at the disc surface layer, high above the mid-plane. Therefore, the comparison between them allows to trace the dynamics present at different vertical heights in the disc.

@@

From previous analysis of radio wavelength observations taken with Sub-millimeter Array (SMA) \citep{Rosenfeld_2013} and subsequently with ALMA \citep{Guzman_2017,Huang_2017,Bergner_2018,Kastner_2018}, it was detected that the millimetre dust configuration of the circumbinary disc had a large inner hole of $\sim$30\,au, featuring a narrow ring centred around 37\,au that accounted for most of the dust mass. By using DPI in the J- and H-band, \citet{Avenhaus_2018} confirmed the double-ring morphology previously presented in \citet{Rapson_2015}, where the observed brightness presented a $\sim$10\,au-wide inner cavity, a narrow ring at $\sim$14\,au and the outer ring centred at $\sim$27\,au.

The suggestion of a parametric model that simultaneously fits the NIR scattered-light and millimetre continuum brightness is challenging and has been attempted in only few previous studies \citep[e.g.][]{Dong_2018,2019MNRAS.486..304B, Ru_z_Rodr_guez_2019}. A successful model can serve as a convenient sample to link the observations to the structure of the underlying dust density distribution.

In this paper we present new observations of the continuum emission at 1.3\,mm of V4046\,Sgr which reveal previously undiscovered structures. This work also aims to reproduce the observations of the protoplanetary disc around V4046\,Sgr with a 3D model using as guidance previous NIR images, the spectral energy distribution (SED) and the new high definition 1.3\,mm continuum ALMA image, as a way of throwing light on its complex structure and the processes that shaped it.

The details of the observations taken into account are described in Section~\ref{sec:Observations}. The parametric model is presented in Section~\ref{sec:model}, the results are discussed in Section~\ref{sec:results} and in Section~\ref{sec:Conclusions} we present our conclusions.

@@

\subsection{SPHERE--IRDIS}  \label{subsec:SPHERE}

@@

The scattered-light image in the top left panel of Fig.~\ref{fig:images_vs_simulated} also shows a double ring structure in the micron-sized dust distribution. The observed morphology presents an inner cavity of $\sim$10\,au in radius and two rings located at 14.10$\pm$0.01\,au, coincident with Ring13, and 24.62$\pm$0.08\,au, coincident with Ring24, with a small gap between them at $\sim$20\,au. The observed second ring matches the inner wall of the 1.3\,mm continuum emission outer ring \citep{Ru_z_Rodr_guez_2019}. Two other important features that are present in the image are: the near-far brightness asymmetry, and the shadows projected on the disc by the close binary system as they eclipse each other, discovered by \citet{dOrazi}.

The binary phase reported by \citet{dOrazi} in the scattered-light observation is at a PA of 265\,$\degr$, east of north. Using this measurement, the binary phase was calculated at the time of the ALMA observation at a PA of $\sim$80\,$\degr$. There is no hint of radio decrements in either Ring13 nor in Ring24 that would match shadowing at this PA. This could be explained by inefficient disc cooling compared to the speed of the illumination pattern \citep[see ][ for estimates of this cooling timescale]{Casassus2019MNRAS.486L..58C}.

@@

The key conclusions of this multi-layered analysis are as follows.
\begin{enumerate}
  \item We report the detection of a two-ringed disc structure. The narrow ring in the 1.3\,mm continuum, has a radius of 13.15$\pm$0.42\,au and an estimated width of 2.46$\pm$0.20\,au. The location of this ring is coincident with the inner ring observed in the scattered-light image. From our RT modeling we can predict that the ring includes around 0.7\,M$_{\earth}$ of millimetre-sized grains. Using the parametric model scale height FWHM value for the large grains ($H=$ 0.25\,au at 13.15\,au) we find that the ring width is roughly 10 times its estimated height. The 1.3\,mm outer ring, that starts at $\sim$24\,au and has its peak intensity at $\sim$30\,au, presents a visible break in the surface brightness at $\sim$36\,au.

@@

  \item Our parametric model, which accounts for the SED of the system, involves the existence of a sub-micron dust population close to the stars (inner disc). Also predicts the presence of another inner ring at $\sim$5\,au, mainly consisting of small dust grains. The depletion of large dust is consistent with efficient dust trapping in Ring13. This additional ring lies under the coronagraph of the scattered-light image and is to faint to be detected by the ALMA observation.
\end{enumerate}

>>>>

% Abstract of the paper
\begin{abstract}
    The nearby V4046\,Sgr spectroscopic binary hosts a gas-rich disc known for its wide cavity and dusty ring. We present high resolution ($\sim$20\,mas or 1.4\,au) ALMA observations of the 1.3\,mm continuum of V4046\,Sgr which, combined with SPHERE--IRDIS polarized images and well-sampled spectral energy distribution (SED), allow us to propose a physical model using radiative transfer (RT) predictions. The new ALMA data reveal a thin ring at a radius of 13.15$\pm$0.42\,au (Ring13), with a radial width of 2.46$\pm$0.20\,au. Ring13 is surrounded by a $\sim$10\,au-wide gap, and it is flanked by a mm-bright outer ring (Ring24) with a sharp inner edge at 24\,au. Between 25 and $\sim$35\,au the brightness of Ring24 is relatively flat and then breaks into a steeper tail that reaches out to $\sim$60\,au. In addition, central emission is detected close to the star which we interpret as a tight ring made of dust grains with a lower size limit of 0.5\,mm. In order to reproduce the SED, the model also requires an inner ring at $\sim$5\,au (Ring5) composed mainly of small dust grains, hiding under the IRDIS coronagraph, and surrounding an inner circumbinary disc. The surprisingly thin Ring13 is nonetheless roughly 10 times wider than its expected vertical extent. The strong near-far disc asymmetry at 1.65\,$\micron$ points at a very forward-scattering phase function, and requires grain radii of no less than 0.4\,$\micron$. The previously reported scattered light shadow of the secondary star is also reproduced by our model.
\end{abstract}

@@

Recent observations of young circumstellar discs have transformed the current knowledge on planet formation. Among the main recent findings of resolved observations of discs is the discovery of substructures in the form of gaps, rings, cavities, and spirals, to name a few \citep[see,][and references there in]{Andrews_2020_aug}. ***EXPAND on rings as radial dust traps***.

The focus of resolved imaging with the Atacama Large Millimeter/submillimeter Array (ALMA) or with the current generation of high-contrast cameras has mainly been towards the brighter sources. It is to address this bias that the ``Discs Around T Tauri Stars with SPHERE'' (DARTTS-S) programme collected differential polarization imaging (DPI) data with the Spectro-Polarimeter High-contrast Exoplanet REsearch \citep[SPHERE][]{2019A&A...631A.155B} for a total of 29 solar-type stars \citep[][]{Avenhaus_2018,Garufi2020}. The sample is not biased towards exceptionally bright and large discs. The DARTTS observations revealed diverse structures and morphologies in the scattering surface of these discs. This article on V4046 Sagittarii (Sgr) is the first installment of a companion programme, the DARTTS survey with ALMA (DARTTS-A), which will present millimetre observations of nine protoplanetary discs previously imaged in polarized scattered light in DARTTS-S.

@@

Previous analysis of radio observations taken with the Sub-Millimeter Array (SMA) \citep{Rosenfeld_2013} and later with ALMA \citep{Guzman_2017,Huang_2017,Bergner_2018,Kastner_2018}, showed that the millimetre dust of the V4046 Sgr circumbinary disc features a large inner hole of $\sim$30\,au and a narrow ring centred around 37\,au that account for most of the dust mass. On the other hand, by using SPHERE polarized light observations in the J- and H-band, \citet{Avenhaus_2018} confirmed the double-ring morphology previously reported by \citet{Rapson_2015} using GPI data. In these polarized light observations, the surface brightness presents a $\sim$10\,au-wide inner cavity, a narrow ring at $\sim$14\,au, and an outer ring centred at $\sim$27\,au.

Finding a parametric model that simultaneously fits the near-IR scattered light and millimetre continuum emission is challenging and has been attempted in a few previous studies \citep[e.g.][]{Dong_2018,2019MNRAS.486..304B, Ru_z_Rodr_guez_2019}. Such model can provide a way to link the observations to the structures in the underlying dust and gas density distributions, and shed light on the complex processes that shaped them.

In this paper, we present new ALMA observations of the continuum emission at 1.3\,mm of V4046\,Sgr which reveal previously undiscovered structures. This work also aims to reproduce these observations with a 3D parametric model that fits all available data, including the polarized scattered light SPHERE image, the spectral energy distribution (SED), as well as the new high definition 1.3\,mm continuum map. Section~\ref{sec:Observations} describes the available observations that we aim to model. Section~\ref{sec:model} describes the structural parameters of our model. A description of our parameter space exploration can be found in Appendix~\ref{sec:Appendix}. The results of our modelling are presented in Section~\ref{sec:results}, including a discussion of the main findings. Finally, in Section~\ref{sec:Conclusions} we present our conclusions.

@@

\subsection{VLT/SPHERE--IRDIS}  \label{subsec:SPHERE}

@@

The scattered light image in the top left panel of Fig.~\ref{fig:images_vs_simulated} also shows a double ring structure in the micron-sized dust distribution. The observed morphology presents an inner cavity of $\sim$10\,au in radius and two rings located at 14.10$\pm$0.01\,au, coincident with Ring13, and 24.62$\pm$0.08\,au, coincident with Ring24, with a small gap between them at $\sim$20\,au. The observed second ring matches the inner wall of the 1.3\,mm continuum emission outer ring \citep{Ru_z_Rodr_guez_2019}. Two other important features that are present in the image are: the near-far brightness asymmetry, and the shadows projected on the disc by the close binary system as they eclipse each other, discovered by \citet{dOrazi}.

The binary phase reported by \citet{dOrazi} in the scattered light observation is at a PA of 265\,$\degr$, east of north. Using this measurement, the binary phase was calculated at the time of the ALMA observation at a PA of $\sim$80\,$\degr$. There is no hint of radio decrements in either Ring13 nor in Ring24 that would match shadowing at this PA. This could be explained by inefficient disc cooling compared to the speed of the illumination pattern \citep[see ][ for estimates of this cooling timescale]{Casassus2019MNRAS.486L..58C}.

@@

The key conclusions of this multi-layered analysis are as follows.
\begin{enumerate}
  \item We report the detection of a two-ringed disc structure. The narrow ring in the 1.3\,mm continuum, has a radius of 13.15$\pm$0.42\,au and an estimated width of 2.46$\pm$0.20\,au. The location of this ring is coincident with the inner ring observed in the scattered light image. From our RT modeling we can predict that the ring includes around 0.7\,M$_{\earth}$ of millimetre-sized grains. Using the parametric model scale height FWHM value for the large grains ($H=$ 0.25\,au at 13.15\,au) we find that the ring width is roughly 10 times its estimated height. The 1.3\,mm outer ring, that starts at $\sim$24\,au and has its peak intensity at $\sim$30\,au, presents a visible break in the surface brightness at $\sim$36\,au.

@@

  \item Our parametric model, which accounts for the SED of the system, involves the existence of a sub-micron dust population close to the stars (inner disc). Also predicts the presence of another inner ring at $\sim$5\,au, mainly consisting of small dust grains. The depletion of large dust is consistent with efficient dust trapping in Ring13. This additional ring lies under the coronagraph of the scattered light image and is to faint to be detected by the ALMA observation.
\end{enumerate}
